---
title: "Getting Started with dagriculture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dagriculture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`dagriculture` is a pure, value-oriented graph library. It is designed to manage graph structures, topological dependencies, structural validity, and task planning. It does not natively orchestrate runtime execution or manage side effects; rather, it tracks structural states (`new`, `ready`, `blocked`) and enforces declarative dependency rules.

## Core Concepts

*   **Registry:** Defines the allowable `kinds` of nodes in a graph.
*   **Graph:** The main immutable data structure carrying nodes, edges, and gates. Operations return a new graph and bump the internal `$version`.
*   **Nodes & Edges:** Represent units of computation (or data) and their dependencies.
*   **Gates:** Explicit structural blockers attached to edges, which prevent downstream nodes from becoming `ready` until the gate is resolved.
*   **Planning:** Topologically sorting and tracking the structural readiness of nodes.

## Installation

```r
pak::pak("sims1253/dagriculture")
```

## Basic Workflow

### 1. Define a Registry

First, establish a registry to strictly define what kind of nodes are allowed in your graphs.

```{r}
library(dagriculture)

reg <- dagriculture_registry(
  dagriculture_kind("input", output_type = "data.frame"),
  dagriculture_kind("model"),
  dagriculture_kind("report")
)
```

### 2. Build the Graph

Start with an empty graph, and add nodes and edges. Because `dagriculture` uses value semantics, every mutating function returns a new updated graph instance.

```{r}
g <- dagriculture_graph(reg)

g <- g |>
  dagriculture_add_node(id = "data", kind = "input", label = "Raw Data") |>
  dagriculture_add_node(id = "fit", kind = "model", label = "Statistical Fit") |>
  dagriculture_add_node(id = "plot", kind = "report", label = "Final Plot") |>
  dagriculture_add_edge(from = "data", to = "fit", id = "e_data_fit") |>
  dagriculture_add_edge(from = "fit", to = "plot", id = "e_fit_plot")

# The internal version tracks changes
g$version
```

### 3. Add a Gate

You can place a "gate" on an edge to represent a manual or external block—for instance, requiring human approval before the statistical fit proceeds.

```{r}
g <- dagriculture_add_gate(g, edge_id = "e_data_fit", id = "approval_gate")
```

### 4. Compute State and Plan

Use `dagriculture_recompute_state()` to evaluate the graph topologically and determine which nodes are structurally ready and which are blocked.

```{r}
g_state <- dagriculture_recompute_state(g)

# The 'data' node is a root and has no blockers
dagriculture_eligible(g_state)

# The 'fit' node is blocked by the gate; the 'plot' node is blocked because 'fit' is blocked.
dagriculture_blocked(g_state)
```

We can generate a structural plan to see how dependencies shake out:

```{r}
plan <- dagriculture_plan(g_state)
plan$targets
plan$blocked
plan$pending_gates
```

### 5. Resolve the Gate

Once the external condition is met, you can resolve the gate. Computing the state again reveals that the downstream dependencies are now unblocked.

```{r}
g_unblocked <- dagriculture_resolve_gate(g_state, id = "approval_gate") |>
  dagriculture_recompute_state()

# Everything is unblocked and ready for execution
dagriculture_eligible(g_unblocked)
dagriculture_blocked(g_unblocked)
```

## Querying Topology

`dagriculture` includes standard helpers for topological traversal and queries.

```{r}
dagriculture_upstream(g_unblocked, "plot")
dagriculture_descendants(g_unblocked, "data")
dagriculture_leaves(g_unblocked)
dagriculture_topo_order(g_unblocked)
```
